Towards typed repositories of proofs

### 2010-06-18 papers/MIPS2010/abstract.tex
déplacement de la partie sur l'ui.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@112 403e655e-980a-47bd-8c60-1ee5348bb6b7

\title{Towards Semantic Patches}
  % pas sûr que le nom soit encore adapté

\author{Matthias Puech, Yann Régis-Gianas}
\date{}

@@

We will describe here some of the possible directions to develop a
tool to analyze the impact of change through types. It involves at its
core a typed description language for repositories -- described in
\S\ref{kernel} -- driven by a user interface that we sketch in
\S\ref{devel}. In the first iteration of this project, we focus on a
static, or data-driven model for repositories; in \S\ref{rw}, we
quickly review related approaches, and hint a possible generalization
in \S\ref{fw}.

@@

Our system will have these properties \emph{wrt.} LF:
\begin{itemize}
\item In this first iteration of the project, we do not need
  computations to take place within our DAGs. Our syntax will then be
  restricted to product types and applications;
\item Every term should be a flat application of variables, so that we
  don't introduce compound terms without naming them;
\item Finally we need a way to record intermediate definitions: we
  introduce a new kind of binder, the equality binder. We maintain the
  invariant that not two same definition can sit in the context while
  typing.
\end{itemize}

@@

Here is a well-typed repository describing the simply-typed
$\lambda$-calculus and constructing two $\lambda$-terms sharing some
sub-derivations: 
{\small
\begin{align}&
(\nat : *).
(o : \nat).
(s : \nat \to \nat).
(type : *).
(base : type).
(arr : type \to type \to type). \\&
(env : *).
(nil : env).
(cons : type \to env \to env). \\&
(term : *).
(var : \nat \to term).
(lam : term \to term).
(app : term \to term \to term). \\&
(lookup : env \to \nat \to type \to *).
(lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
(lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
(of : env \to term \to type \to *). \\&
(of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
          lookup\ e\ n\ a \to of\ e\ v\ a). \\&
(of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
          (u = lam\ t : term).(f = cons\ a\ e : env).
          of\ f\ t\ b \to of\ e\ u\ c). \\&
(of_a : (e : env).(t\ u: term).(a\ b: type).
          (c = arr\ a\ b : type).(v = app\ t\ u : term).
          of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
(1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
(t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
(t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
(t_4 = lam\ t_2 : term).
(bb = arr\ base\ base : type). \\&
(e_0 = cons\ bb\ nil : env).
(e_1 = cons\ base\ e0 : env).
(d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
(d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
(d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
(bbb = arr\ bb\ bb : type).
(d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
(g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
(g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
(g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
(t_5 = app\ v_1\ v0 : term). \\&
(g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
(t_6 = lam\ t_5 : term). \\&
(g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
(t_7 = lam\ t_6 : term).
(g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
\end{align}
}

\section{Development through semantic patches}
\label{devel}

\remplan{Fri Jun 18, 2010 10:50 AM. L'objectif ici, c'est de décrire
  une (possible) architecture pour rendre utilisable le langage 
  décrit dans la section précédente. En gros, on énumère les 
  opérations qui interviennent dans l'utilisation et on les décrit
  informellement à l'aide d'un exemple.}

The kernel language described in the previous section is clearly not
adapted to human interaction, for several reasons~: 
\begin{itemize}

\item Every single object is flattened, which makes unobvious the
  structure of a program or a proof.

\item A term of this language denotes the entire history of a
  development. Most of the time, the user is more interested in a
  instantaneous snapshot of his object of study.

\item Even though we were able to provide a snapshot of a particular
  state of the development, the repository represents this state 
  using proof terms, which contain too many syntactic details from the
  user point of view and are not suitable for edition.

\end{itemize}

\remtext{Fri Jun 18, 2010 10:52 AM. Par contre, je ne sais pas quel
  exemple utilisé. Le STLC? Est-ce que ça fait ``assez'' système
  logique pour ce workshop?}

\subsection{A layered architecture}

\remtext{Fri Jun 18, 2010  1:32 PM. Ici, on décrit la figure en suivant
  le modèle de flot du système. On décrit le rôle des fonctions, 
les problèmes posés.}

% À mettre dans une figure. 
\begin{figure}
\begin{center}
% FIXME: il y a un problème avec la flèche double phi/psi.
\xymatrix{
\textrm{History of metatheory objects}
&R 
\ar@{->}[d]_{\phi}
& \Delta \ar@{->}[r]^{\theta(R)} & R' \\
\textrm{Metatheory objects} 
&t \ar@{->}[d]^{\pi}
\ar@{->}[u]_{\psi}
& t' \ar@{->}[u]_{\nabla(t)}\\
\textrm{Concrete views on metatheory objects} 
&w \ar@{.>}[r] & w' \ar@{->}[u]_{\sigma(t, w)}
}
\end{center}
\label{fig:architecture}
\caption{A layered architecture for a semantic patch framework.}
\end{figure}

With the previous remarks in mind, we designed a generic framework of semantic
patches that can be described using three layers which are summarized in 
Figure~\ref{fig:architecture}. We now describe the role of each layer.

The most internal layer focused on the history of metatheory
objects. It is composed of a type-checker for the kernel language. The
purpose of this layer is to maintain a repository~$R$ and the
invariant that it is well-typed. Through the layer interface, a client
can ask for the \textit{integration} $\theta(R)$ of a patch $\Delta$
to obtain an extended version $R'$ of the repository $R$.

The middle layer is a mechanized formalization of a metatheory. This
module provides a syntax for the objects of the metatheory (formulae,
judgments, \ldots) as well as the admissible rules (that we also call
metatheorems).  An \textit{internalization} function $\phi$ translates
the constructors of these objects and the specification of the
admissible rules as terms of the kernel language. As a consequence,
every object~$t$ of the theory can be injected into the
repository~$R$. An \textit{externalization} function $\psi$ is
implementing the inverse operation. Given a modified version $t'$ of
an object~$t$, we are able to compute the patch $\Delta$ using
the \textit{differentiation} function $\nabla(t)$.

The external layer offers views on the objects of the theory. A
view~$w$ is a human readable, and editable, partial representation of
an object~$t$. A view is obtained from an object through a
projection~$\pi$.  The user transforms the view $w$ into a view $w'$
which must be $\textit{put back}$ into the metatheory as a modified
object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.  In
concrete syntax, an object~$t$ may be denoted simply using its
syntactic definition but also, more interestingly, by a reference to this 
object into the repository.

The implementation of this architecture raises some interesting
technical difficulties that can be handled, but only partially, 
by existing technology. Indeed, even if internalization and 
externalization are easily implemented thanks to the expressivity
of the kernel language, the projection, putback, differentiation
and integration operations are more intricate to realize.

\paragraph{From an abstract proof tree to a concrete view, and back}
\-
\remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
  pas très pratique. Pourtant, c'est moralement sur cet objet 
  que s'opère le développement et non sur le code source, qui
  n'en est qu'une représentation. On justifie la nécessité de
  projection et on fait référence aux lentilles de Pierce pour
  expliquer $\sigma$ et $\pi$.}

The problem of updating some data according to a modification that was
done on a partial view of it is a well-known issue in database
literature called the \textit{view-update} problem. Pierce \textit{et
  al} have recently attacked it from a language design approach. Their
idea is to develop projection and putback functions simultaneously by
writing a so-called lens in a bidirectional programming
language~\cite{pierce-bidirectional}. Roughly speaking, a lens is a
software component that can be used to produce output from input but
also input from output (hence the bidirectionality). A lens is able
to keep track of the information that was erased by a projection to
produce a projected output from the initial input. When we supply a
modified output to the lens, it can elaborate, from the initial input
and under certain conditions, the missing part of the modified input.

We plan to design a bidirectional language to define the
projection~function~$\pi$ and the putback~function~$\sigma$ as a
lens. Yet, we cannot just import existing lenses combinators from the
Harmony system that already works on trees. Indeed, most of the time,
our trees have extra structural constraints induced by the presence of
binders.

\remtext{Fri Jun 18, 2010 1:46 PM. Autre point important sur les vues
  concrètes~: elles peuvent servir à parler de l'historique.  Voir le
  module MetaAST de l'implémentation. Ces vues sont utiles par exemple
  lorsque l'utilisateur doit construire un patch d'adaptation. Il a
  alors toute la puissance des métathéorèmes ainsi que tous les objets
  déjà construits pour décrire son patch.}

\paragraph{Capturing the evolution of abstract proof trees}
\-

Computing the minimal edition script between two abstract syntax trees
is a NP-hard problem that has been heavily studied, in particular in
bio-informatics.  We guess that, in practice, we can restrict the
search to local transformations that preserve substructures (that
roughly consist in inserting, removing or moving around entire
sub-trees). In that case, an existing polynomial algorithm may be
used~\cite{Touzet05alinear}. Besides, abstract trees are relatively small in
practice.

We are more concerned about the fact that existing algorithms are not
aware of binders that occur in proof-terms. Indeed, some standard
edition scripts must be forbidden because they could break
well-scopedness. As a consequence, we will have to adapt standard
algorithms to only look for scope-preserving edition scripts.

%% We are more concerned about our ability to elaborate high-level patches
%% from these low-level edition scripts. Actually, it is easy to deduce
%% from a low-level edition script, a low-level patch, based on object
%% syntax constructor elimination and introduction. However, such a
%% low-level patch may forbid a more efficient reuse of existing
%% objects. For instance, let us assume a transformation of an
%% object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
%% would represents the transformation as the replacement of the first
%% sub-term $x$ by the sub-term~$2$ followed by the replacement of the
%% second sub-term $x$ by the sub-term~$2$. If we had an object-level type
%% derivation for $x + x$, the integration function would have to build
%% two sub-derivations, one for each replaced sub-terms.  Instead, a
%% high-level patch would represent that transformation as a substitution
%% of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
%% that case, the integration function has the opportunity to apply a
%% metatheorem about the stability of typing derivations under
%% substitution and finally reuse the initial typing derivation
%% directly. Inferring automatically such a high-level patch seems an
%% unreachable objective~: there is clearly no most general high-level
%% patch. Nonetheless, an interactive process can be designed to help the
%% user building his high-level patch from a set of pre-computed
%% interpretations of the low-level patch.

\remtext{Fri Jun 18, 2010 1:38 PM. Il faut différencier deux arbres de
  syntaxe. Une méthode : la distance d'édition syntaxique permet de
  déduire des applications des règles d'élimination et d'introduction
  que l'on peut utiliser pour construire $\Delta$. Référence sur la
  littérature sur les distances d'édition entre arbres. Discussion sur
  l'élaboration de patchs de haut niveau (comme le renommage etc). }

\paragraph{Propagating changes through dependencies}
\-

There is several ways of integrating a patch $\Delta$ inside a
repository~$R$.  The naive attempt would be a simple concatenation
of~$\Delta$ at the end of~$R$.  This kind of integration, though
possible and correct, is of little interest because it does not take
into account the fact that $\Delta$ probably defines new versions of
objects already present in~$R$.

A more useful integration propagates changes to the context that uses
the modified object in order to trigger the updating of this context
(which may trigger the updating of other tree, and so forth). Again,
in full generality, it is not possible to update the context
automatically. Instead, we choose to generate \textit{challenges},
which are adaptation patches with holes. These challenges can be
addressed by the author that wrote the patch to be integrated, by the
author of the context or automatically when there is no ambiguity
(think of a context that uses a function whose body has changed but
not its type).

\remtext{Fri Jun 18, 2010 1:42 PM. Si on se contente de concaténer
  $\Delta$ à $R$, on n'a intégré qu'une nouvelle version de l'objet
  initial~$t$ dans $R$ mais on n'a pas propagé ces modifications
  auprès des clients de $t$. Cette propagation n'est \textit{a priori}
  pas triviale~: rien ne dit que $t'$ à un sens équivalent à $t$. Il
  faut donc adapter les clients en demandant un patch d'adaptation à
  l'utilisateur. On présente ces demandes sous la forme de patch à
  compléter (on fournit un peu de contexte et on a un trou pour y
  injecter une nouvelle sous-dérivation mais on peut aussi modifier le
  contexte pour l'adapter). La construction d'un patch d'adaptation
  est un processus interactif~: on ne sait pas a priori borner le
  contexte nécessaire à la construction du patchs d'adaptation donc on
  peut imaginer un sorte de jeu où le système fait augmenter le
  contexte (en suivant les dépendances) si l'utilisateur ne sait pas
  construire le patch d'adaptation. En passant, quand on parle
  d'utilisateur ici, ce peut être aussi bien un programme, comme un
  type-checker pour le langage objet, ou bien un programmeur.}

\subsection{Application on simply typed $\lambda$-calculus}

@@

%% We propose here to devise a system for \emph{semantic patches}. It
%% % TODO changer le nom?
%% substitutes the idea of textual transformation by:
%% \begin{itemize}
%% \item First preferring an abstract syntax tree representation instead
%%   of plain text;
%% \item Secondly embedding semantic, or typing data into the
%%   transformations, in order to be able to reason on them.
%% \item Finally, perform type-checking in an incremental way, that is
%%   type only the syntactic difference and reuse derivation for the
%%   already known subprograms.
%% \end{itemize}

%% We want to design a \emph{language of patches}, able to represent
%% % TODO changer le nom?
%% these transformations and their semantic properties, thus capturing
%% local syntactic changes in programs (the abstract syntax tree), as
%% well as their global effect on the whole project (the typing
%% derivations). We can see this goal as a refinement of the former idea
%% of ``dependency''.

>>>>

\title{Towards typed repositories of proofs}

\author{Matthias Puech \and Yann Régis-Gianas}
\date{}

@@

We will describe here some of the possible directions to develop a
tool to analyze the impact of change through types. It involves at its
core a typed description language for repositories, described in
\S\ref{kernel}. In the first iteration of this project, we focus on a
static, or data-driven model for repositories; in \S\ref{rw}, we
quickly review related approaches, and hint a possible generalization
in \S\ref{fw}.

@@

Our system will have these properties w.r.t. LF:
\begin{itemize}
\item In this first iteration of the project, we do not need
  computations to take place within our DAGs. Our syntax will then be
  restricted to product types and applications;
\item Every term should be a flat application of variables, so that we
  don't introduce compound terms without naming them;
\item Finally we need a way to record intermediate definitions: we
  introduce a new kind of binder, the equality binder. We maintain the
  invariant that not two same definition can sit in the context while
  typing.
\end{itemize}

@@

Here is a well-typed repository describing the simply-typed
$\lambda$-calculus and constructing two $\lambda$-terms sharing some
sub-derivations: 
{\small
\begin{align}&
(\nat : *).
(o : \nat).
(s : \nat \to \nat).
(type : *).
(base : type).
(arr : type \to type \to type). \\&
(env : *).
(nil : env).
(cons : type \to env \to env). \\&
(term : *).
(var : \nat \to term).
(lam : term \to term).
(app : term \to term \to term). \\&
(lookup : env \to \nat \to type \to *).
(lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
(lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
(of : env \to term \to type \to *). \\&
(of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
          lookup\ e\ n\ a \to of\ e\ v\ a). \\&
(of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
          (u = lam\ t : term).(f = cons\ a\ e : env).
          of\ f\ t\ b \to of\ e\ u\ c). \\&
(of_a : (e : env).(t\ u: term).(a\ b: type).
          (c = arr\ a\ b : type).(v = app\ t\ u : term).
          of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
(1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
(t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
(t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
(t_4 = lam\ t_2 : term).
(bb = arr\ base\ base : type). \\&
(e_0 = cons\ bb\ nil : env).
(e_1 = cons\ base\ e0 : env).
(d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
(d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
(d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
(bbb = arr\ bb\ bb : type).
(d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
(g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
(g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
(g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
(t_5 = app\ v_1\ v0 : term). \\&
(g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
(t_6 = lam\ t_5 : term). \\&
(g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
(t_7 = lam\ t_6 : term). \\&
(g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
\end{align}
}

@@

%% We propose here to devise a system for \emph{semantic repositories}. It
%% % TODO changer le nom?
%% substitutes the idea of textual transformation by:
%% \begin{itemize}
%% \item First preferring an abstract syntax tree representation instead
%%   of plain text;
%% \item Secondly embedding semantic, or typing data into the
%%   transformations, in order to be able to reason on them.
%% \item Finally, perform type-checking in an incremental way, that is
%%   type only the syntactic difference and reuse derivation for the
%%   already known subprograms.
%% \end{itemize}

%% We want to design a \emph{language of repositories}, able to represent
%% % TODO changer le nom?
%% these transformations and their semantic properties, thus capturing
%% local syntactic changes in programs (the abstract syntax tree), as
%% well as their global effect on the whole project (the typing
%% derivations). We can see this goal as a refinement of the former idea
%% of ``dependency''.
